TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2019-06-10 paper/draft.tex
make doc more readable

Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citep{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.

@@

A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g. \citealt{Terrien:2012}, \citealt{Rojas-Ayala:2012}, \citealt{Newton:2014}).
However these metallicity calibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant calibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.

In this work we build a data-driven model for M dwarf spectra with \thecannon\ (\citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}, \citealt{Behmard2019}).
\thecannon\ is a pixel-by-pixel regression for spectral data given a training
set of stars with reliable (assumed) accurate labels.
The model can both predict a spectrum given labels, and also predict labels given a spectrum.
Thus it is a generative model for stellar spectra that can be used for parameter inference.

@@

The \apogee\ survey 
% \citep{Majewski:2015} of the Sloan Digital Sky Survey III and IV (\textsl{SDSS-III/IV} \citealt{Eisenstein:2011}; \citealt{Blanton:2017})
is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}). For low temperatures ($2800<\teff<3500$\,K), the pipeline uses \textsl{MARCS} plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) \textsl{ATLAS9} \citep{Castelli:2004} plane-parallel models are used.

@@

\subsection{Training Sample Selection}

@@

\subsection{Temperature/Metallicity Model \label{subsec:mann_results}}
For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex. We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr for the published sample estimated by M15) and likely slowly rotating sources. While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.

@@

\subsection{Spectral Type Model \label{subsec:west_results}}
We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type. We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source. As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space. Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows the self and LOOCV tests for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.

@@

\subsection{Color-Temperature Comparisons}

@@

\subsection{Isochrone Comparisons}

@@

\begin{figure*}[ht]
% \plotone{figures/self_and_validation_test_teff_snr.png}
% \plotone{figures/self_and_validation_test_feh_snr.png}
\plottwo{figures/crv_test_teff.png}{figures/crv_test_fe_h.png}
\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right). Sources with strong rotational broadening (2M07171706--0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles. Predictive accuracy, as computed by the scatter in cross validation is 77K in temperature, and 0.09 dex in metallicity.}\label{fig:mann_validation}
\end{figure*}

\begin{figure*}[ht]
\begin{center}
% \plotone{figures/self_and_validation_test_spt_snr.png}
\plottwo{figures/self_test_spt.png}{figures/crv_test_spt.png}
\end{center}
\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).  Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.}\label{fig:west_validation}
\end{figure*}

>>>>

Advances in instrumentation with the lauch of several spectroscopic surveys in the past decade, such as the Sloan Digital Sky Survey (SDSS), and the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST) have dramatically increased the sample of known M dwarfs (\citealt{West:2011}, \citealt{Guo:2015} with catalogs of over 70,000 sources) and our potential to study M dwarf abundances on both exoplanet and galactic scales at a large statistical level.
The launch of the Apache Point Observatory Galactic Evolution Experiment (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III/IV mission \citep{Eisenstein:2011,Blanton:2017} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy \citep{Desphande:2013}, yet pipeline measurements \citep{Perez:2016} of $\teff$ and $\feh$ have only been determined precise to 100\,K and 0.18\,dex down to $T_{\rm eff}\sim3550\,$K using atmospheric models \citep{Schmidt:2016}.

@@

A number of studies focused on improving M dwarf metallicity precisions have used systems of M dwarfs in common proper motion with an FGK star, to developed precise empirical relations (as good as $\sim$0.07\,dex) between the adopted metallicities of the FGK primary (under the assumption kinematically associated wide binaries are co-evolutionary; \citealt{Bonfils:2005}), and strong, isolated lines in the optical spectra of the M dwarf (e.g. \citealt{Terrien:2012}; \citealt{Rojas-Ayala:2012}; \citealt{Newton:2014}).
However these metallicity calibrations are inaccesible for analysis with \apogee, as the narrow infrared wavelength coverage does not include the relevant calibration indices used by existing studies (such as Ca I or KI lines in the optical), and blending of multiple atomic and molecular species at high resolution makes it impractical to identify and deconvolve new individual features for analysis.

In this work we build a data-driven model for M dwarf spectra with \thecannon\ \citep{Ness:2015,Casey:2016,Ho:2017a}.
\thecannon\ is a pixel-by-pixel regression for spectral data given a training
set of stars with reliable (assumed) accurate labels.
The model can both predict a spectrum given labels, and also predict labels given a spectrum.
Thus it is a generative model for stellar spectra that can be used for parameter inference.

@@

The \apogee\ survey is a high resolution (R$\sim$22,500), H-band (1.5-1.7$\mu$m), multi-epoch survey which has observed over 250,000 stellar spectra up to its fourteenth data release (DR14; \citealt{Abolfathi:2017}). Fundamental parameters for each of these stars are estimated by the \apogee\ Stellar Parameter and Chemical Abundances Pipeline (\aspcap; \citealt{Perez:2016}), which employs a $\chi^2$ fitting procedure using the \textsl{FERRE} code to fit radiative transfer models and determine atmospheric parameters, 15 chemical abundances and micro-turbulence parameters ($\teff$, $\logg$, $\feh$, [$\alpha$/M], [C/M], [N/M]; \citealt{Meszaros:2012}). For low temperatures ($2800<\teff<3500$\,K), the pipeline uses \textsl{MARCS} plane-parallel/spherical models \citep{Gustafsson:2008}, and for higher temperatures ($\teff\geq3500$\,K) \textsl{ATLAS9} \citep{Castelli:2004} plane-parallel models are used.

@@

% ----------------------------------------
\subsection{Training Sample Selection}

@@

\subsection{Temperature/Metallicity Model \label{subsec:mann_results}}

For the physical parameter model we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex. We note for this model 6 out of 87 sources show strong visible rotational line broadening (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources are mostly old ($>8$ Gyr for the published sample estimated by M15) and likely slowly rotating sources. While these sources with high rotational broadening have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), we find that the scatter of labels derived for these sources is consistent with the rest of the sample and we do not remove them from the training sample.

@@

% ----------------------------------------
\subsection{Spectral Type Model \label{subsec:west_results}}

We trained \thecannon\ on 51 M dwarfs with a one-dimensional spectral type label, and obtained a precision of $\pm$0.9 spectral types, also similar to the original training label uncertainty of $\pm$1 spectral type. We note however, that the training sample is distributed heavily towards earlier type sources, with a median spectral type of 3 and only one M8 and one M9 source. As seen in Figure \ref{fig:west_validation}, the model performs well at re-deriving the labels for the M8 and M9 source when they are included in the model (less than 1 SPT difference), but badly when left out (over 2 SPT difference), which indicates \thecannon\ does a poor job at extrapolating to labels outside of the training sample space. Repeating the analysis of Section \ref{subsec:mann_results}, Figure \ref{fig:west_validation} shows the self and LOOCV tests for the labels reported in Table \ref{table:west_results}, and Figure \ref{fig:west_derivative} shows the derivative of model flux with varying spectral type.

@@

% ----------------------------------------
\subsection{Color-Temperature Comparisons}

@@

% ----------------------------------------
\subsection{Isochrone Comparisons}

Figure \ref{fig:isochrone_cmd}

@@

\begin{figure*}[ht]
	\plottwo{figures/crv_test_teff.png}{figures/crv_test_fe_h.png}
	\caption{Consistency test for the Mann-trained temperature/metallicity type model; self-test (left), and leave-one-out cross validation (right). Sources with strong rotational broadening (2M07171706--0501031, 2M08155393+3136392, 2M09174473+4612246, 2M10331367+3409120, 2M15594729+4403595, 2M13400879+4346380), as identified by visual inspection are marked with red circles. Predictive accuracy, as computed by the scatter in cross validation is 77K in temperature, and 0.09 dex in metallicity.}\label{fig:mann_validation}
\end{figure*}

\begin{figure*}[ht]
	\begin{center}
	\plottwo{figures/self_test_spt.png}{figures/crv_test_spt.png}
	\end{center}
	\caption{Consistency test for the West-trained spectral type model; self-test (left), and leave-one-out cross validation (right).  Predictive accuracy, as computed by the scatter in cross validation is 0.9 subtypes.}\label{fig:west_validation}
\end{figure*}
